---
title: "Why We Need Bayesian Linear Regression"
subtitle: "Mechanism, Interpretation, and Failure Modes of Frequentist OLS"
author: "Daniel Skak Mazhari-Jensen"
date: today
format:
  revealjs:
    theme: moon
    slide-number: true
    code-fold: true
    scrollable: true
---

# What is this Bayesian thing anyway?

Chat with your neighbor for 3 minutes - where/what have you heard about Bayesian inference?

# :raising_hand_woman: How many have heard about Bayesian inference? {.center}

# :raising_hand_woman: How many already use a Bayesian workflow? {.center}

# :walking::walking_woman: Do you ????? {.center}

Never <------->
I took a course in computer science

# The Core Question {.smaller}

Why do we need a Bayesian approach to regression - we can just use ordinary least squares?

The answer is not (only):

-  philosophy
-  elegance
-  a matter of preference

::: {.callout-important appearance="simple" icon=false}
Real regression problems are often weakly identified, noisy, unstable, and geometrically pathological.
:::

Bayesian regression stabilizes inference.

# The Frequentist Ideal {.smaller}

Ordinary Least Squares (OLS):

-  minimizes squared residuals
-  produces unbiased estimators
-  provides confidence intervals and *p*-values
-  works beautifully asymptotically

The classical linear model:

$$
y = X\beta + \epsilon
$$

with

$$
\epsilon \sim \mathcal{N}(0, \sigma^2)
$$

OLS estimator:

$$
\hat{\beta}_{OLS} = (X^TX)^{-1}X^Ty
$$

# The Hidden Assumption

OLS works well when:

-  predictors are not highly collinear
-  sample size is large
-  signal-to-noise ratio is reasonable
-  parameters are strongly identified

But real data often violate all of these simultaneously.

# A More Realistic Situation {.smaller}

Suppose we have:

-  30 observations
-  10 predictors
-  strong predictor correlation
-  weak signal
-  uncertain mechanisms

This is extremely common in:

-  social science
-  biology
-  psychology
-  economics
-  policy research

# The Data Generating Process {.smaller}

The truth is simple.

Only one predictor matters.

$$
y = 1.5x_1 + \epsilon
$$

All other predictors have zero effect.

Yet predictors are highly correlated.

# The Mechanism of Failure {.smaller}

With correlated predictors:

-  many coefficient combinations fit equally well
-  the likelihood becomes nearly flat
-  tiny noise changes estimates dramatically

OLS asks:

::: {.callout-note appearance="simple" icon=false}
Which coefficients minimize prediction error?
:::

But it never asks:

::: {.callout-warning appearance="simple" icon=false}
Are these coefficient magnitudes plausible?
:::


# The Geometry Problem

Collinearity creates unstable directions in parameter space.

The matrix:

$$
(X^TX)^{-1}
$$

becomes nearly singular.

As a result:

-  coefficients explode
-  signs flip
-  uncertainty becomes enormous
-  interpretation collapses



# The Bayesian Solution

Bayesian regression adds prior information:

$$
\beta_j \sim \mathcal{N}(0,1)
$$

Posterior:

$$
p(\beta \mid y)
\propto
p(y \mid \beta)p(\beta)
$$

The prior regularizes weakly identified directions.

This changes the geometry of inference.



# Important Clarification

Bayesian priors are not magic.

They encode structural skepticism.

The prior says:

::: {.callout-tip appearance="simple" icon=false}
Large coefficients should require strong evidence.
:::

This is often scientifically reasonable.



# Simulation Setup

We now simulate the failure directly.

-  small sample
-  many predictors
-  high collinearity
-  weak signal

The true model:

$$
y = 1.5x_1 + \epsilon
$$

Only x1 matters.



# Simulating Correlated Predictors

```{r modeling}
#| echo: false
#| output: false
library(MASS)
library(corrplot)

set.seed(12)

n <- 30
p <- 10

Sigma <- matrix(0.9, p, p)
diag(Sigma) <- 1

X <- mvrnorm(n, mu = rep(0, p), Sigma = Sigma)

colnames(X) <- paste0("x", 1:p)

# Generating the Outcome
beta_true <- c(1.5, rep(0, 9))

y <- X %*% beta_true + rnorm(n, 0, 2)

dat <- data.frame(y, X)

# Freq
fit_lm <- lm(y ~ ., data = dat)

# Bayes
library(rstanarm)

fit_bayes <- stan_glm(
  y ~ .,
  data = dat,
  prior = normal(0, 1),
  prior_intercept = normal(0, 5),
  chains = 4,
  iter = 2000,
  refresh = 0
)
```

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation structure among predictors."
corrplot(cor(X), method = "color")

head(dat)
```


# Frequentist OLS

```{r}
#| label: tbl-ols
summary(fit_lm)
```

# Why This Happens

Predictors are interchangeable.

The model can fit equally well using:

-   huge positive x1
-   huge negative x2

OLS has no mechanism preventing absurd parameter values.

It only optimizes fit.

# The Frequentist Interpretation Problem

Researchers now face impossible interpretation.

Questions become unstable:

-   Which variables matter?
-   Which signs are trustworthy?
-   Which effects are real?

*p*-values fluctuate wildly across samples.


# Bayesian Regression

```{r}
#| label: bayes-fit
#| 
print(fit_bayes, digits = 2)
```

# Let's have a look 

```{r}
#| label: model-plot
#| fig-cap: "model estimates and ground truth."
bayes_draws <- fit_bayes |>
  #tidybayes::spread_draws(`(Intercept)`, `^x\\d+`, regex = TRUE) |>
  tidybayes::spread_draws(`^x\\d+`, regex = TRUE) |>
  tidyr::pivot_longer(
    #cols = c(`(Intercept)`, tidyselect::starts_with("x")),
    cols = tidyselect::starts_with("x"),
    names_to = "term",
    values_to = "estimate"
  ) |>
  dplyr::mutate(model = "Bayesian")


freq_df <- broom::tidy(fit_lm, conf.int = TRUE) |>
  dplyr::select(term, estimate, conf.low, conf.high) |>
  dplyr::mutate(model = "Frequentist")

truth_df <- tibble::tibble(
  term = c("(Intercept)", paste0("x", 1:p)),
  truth = c(0, beta_true),
  model = "Truth"
)

bayes_summary <- bayes_draws |>
  dplyr::group_by(term) |>
  tidybayes::median_qi(estimate, .width = 0.95) |>
  dplyr::transmute(
    term,
    estimate,
    conf.low = .lower,
    conf.high = .upper,
    model = "Bayesian"
  )


freq_plot <- freq_df |>
  dplyr::transmute(
    term,
    estimate,
    conf.low,
    conf.high,
    model = "Frequentist"
  )

plot_df <- dplyr::bind_rows(
  bayes_summary,
  freq_plot
)

ggplot2::ggplot(
  plot_df,
  ggplot2::aes(
    x = estimate,
    y = term,
    color = model
  )
) +

  ggdist::geom_pointinterval(
    ggplot2::aes(
      xmin = conf.low,
      xmax = conf.high
    ),
    position = ggplot2::position_dodge(width = 0.5)
  ) +

  ggplot2::geom_point(
    data = truth_df,
    ggplot2::aes(x = truth, y = term),
    inherit.aes = FALSE,
    color = "blue",
    shape = 4,
    size = 3,
    stroke = 1.2
  ) +

  ggplot2::geom_vline(
    xintercept = 0,
    linetype = 2
  ) +

  ggplot2::theme_bw()
```

# What the Prior Does

The prior shrinks implausible coefficients toward zero.

Not aggressively.

Just enough to stabilize weakly identified directions.

This is called:

-   regularization
-   shrinkage
-   partial pooling of uncertainty

# Bias-Variance Tradeoff

Frequentist OLS emphasizes unbiasedness.

But unbiased estimators can have enormous variance.

Bayesian regression accepts:

-   tiny bias
-   dramatically lower variance

This often improves:

-   prediction
-   calibration
-   scientific interpretation

# The Deep Point

OLS treats the model as fixed truth.

Bayesian workflow treats models as uncertain approximations.

This distinction is profound.

# Posterior Interpretation

Frequentist confidence interval:

::: {.callout-warning appearance="simple" icon=false}
Does NOT mean:
"There is a 95% probability the parameter lies here."
:::

Bayesian posterior interval:

::: {.callout-important appearance="simple" icon=false}
Literally means:
"Given model and data, there is 95% posterior probability the parameter lies here."
:::

# Weak Identification

The Bayesian posterior honestly expresses:

-   uncertainty
-   partial identifiability
-   lack of information

Sometimes the correct scientific answer is:

::: {.callout-note appearance="simple" icon=false}
"We do not know very much."
:::

OLS often hides this instability behind noisy point estimates.

# Posterior Predictive Thinking

Bayesian models naturally support:

-   posterior predictive checks
-   hierarchical models
-   measurement error
-   causal modeling
-   uncertainty propagation

This enables model criticism rather than blind estimation.

# A More Extreme Failure

Increase:

-   predictors from 10 to 20
-   correlation from 0.9 to 0.95

OLS becomes nearly meaningless.

Coefficients:

-   explode
-   reverse sign
-   become sample-dependent noise

Bayesian regularization still produces stable inference.

# The Core Mechanism

Frequentist OLS:

$hat{β}$ =arg min RSS

Bayesian regression:

$$
p(\beta \mid y)
\propto
p(y \mid \beta)p(\beta)
$$

The prior changes the geometry of inference.

This is the key idea.

# Final Takeaway

Bayesian regression matters because:

-   real models are weakly identified
-   finite samples are noisy
-   predictors are correlated
-   unconstrained likelihoods become pathological

Priors repair inferential geometry.

# Final Thought

The Bayesian question is not:

::: {.callout-note appearance="simple" icon=false}
"What coefficient minimizes error?"
:::

The Bayesian question is:

::: {.callout-important appearance="simple" icon=false}
"What parameter values remain plausible after combining data with scientific structure?"
:::

That is usually the more meaningful scientific question.

# References {#sec-references}

Gelman, A., et al. Bayesian Data Analysis
McElreath, R. Statistical Rethinking
Vehtari, A., Gelman, A., Gabry, J. (2017)
"Practical Bayesian model evaluation using leave-one-out cross-validation"
Hastie, Tibshirani, Friedman. Elements of Statistical Learning