---
title: "Why be a Bayesian"
subtitle: "A non-technical introduction to the Bayesian approach"
author: "Daniel S. Mazhari-Jensen"
description: "Bayesian inference is not only elegant and intuitive, but also helpful for overcoming common challenges with data from the life sciences"
date: 2026-05-26
reading-time: 10 min
file-modified: 2026-05-26
categories: [Bayesian, AI life science, course material]
format:
  html:
    toc: true
    code-fold: true
---

> [**Slides**](../_site/ideas_for_future_blog_posts/why_we_bayesian_slides.html)

<div>

```{=html}
<iframe class="slide-deck" src="../slides/why_we_bayesian_slides.html"></iframe>
```

</div>

::: {.callout-note collapse="true"}
## Teacher note

The slides contain speaking notes that you can view by pressing 'S' on the
keyboard.
:::

# Why be a Bayesian?

There are many reasons to be a Bayesian.

# What is the alternative?

Frequentist statistics

# Is there a fight?

No! That fight ended in the 90's

# When to use Frequentist and when to use Bayesian approaches?

Frequentist are often sufficient and comparable to Bayesian stats.

# But sometimes, they are very different!


```{r make-data}
#| output: false
#| echo: false

# install.packages(c("MASS", "rstanarm", "tidyverse"))

library(MASS)
library(rstanarm)
library(tidyverse)

set.seed(18062026)

# -----------------------------
# Simulate correlated predictors
# -----------------------------

p <- 30 # different bio-sample tests (30 genes)
n <- 32 # Participants/samples

Sigma <- matrix(0.8, p, p) # Make covar struct
diag(Sigma) <- 1 # and var

X <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = Sigma) # var-covar struct

colnames(X) <- paste0("x", 1:p) 

# True data-generating process:
# only x1 matters

beta_true <- c(2, rep(0, p - 1)) # Make true effect

y <- X %*% beta_true + rnorm(n, 0, 2) # Ingest true effect and gaussian noise

dat <- data.frame(y, X, id = 1:n) # Make data.frame
dat <- dat |>
  dplyr::mutate(
    id = as.factor(id)
  )

dat_long <- dat |>
  tidyr::pivot_longer(
    cols = starts_with("x"),
    names_to = "predictor",
    values_to = "value",
  )

form <- as.formula(
  paste("y ~", paste0("x", 1:p, collapse = " + "))
)
```


```{r freq-model}
#| output: false
#| echo: false

# -----------------------------
# Frequentist OLS
# -----------------------------

#fit_lm <- lm(y ~ x1, data = dat) # lm() expects wide format!
fit_lm <- lm(form, data = dat)

summary(fit_lm)
```


```{r bayes-model}
#| output: false
#| echo: false

# -----------------------------
# Bayesian regression
# -----------------------------

fit_bayes <- rstanarm::stan_glm(
  form,
  data = dat,
  prior = rstanarm::normal(0, 1.5),
  prior_intercept = rstanarm::normal(0, 0.25),
  chains = 4,
  iter = 2000,
  refresh = 0
)

print(fit_bayes, digits = 2)
```


```{r prep-plot}
#| output: false
#| echo: false


fit_bayes |>
  tidybayes::spread_draws(
    #`(Intercept)`,
    `^x\\d+`,
    regex = TRUE
  ) |>
  #tidybayes::median_qi(condition_mean = `(Intercept)` + b, .width = c(.95, .66)) |>
  tidybayes::median_qi(x1, .width = c(.95, .66)) |>
  #ggplot2::ggplot(aes(y = group, x = condition_mean, xmin = .lower, xmax = .upper)) +
  ggplot2::ggplot(aes(y = x1, x = x1, xmin = .lower, xmax = .upper)) +
  ggdist::geom_pointinterval()


####

bayes_draws <- fit_bayes |>
  #tidybayes::spread_draws(`(Intercept)`, `^x\\d+`, regex = TRUE) |>
  tidybayes::spread_draws(`^x\\d+`, regex = TRUE) |>
  tidyr::pivot_longer(
    #cols = c(`(Intercept)`, tidyselect::starts_with("x")),
    cols = tidyselect::starts_with("x"),
    names_to = "term",
    values_to = "estimate"
  ) |>
  dplyr::mutate(model = "Bayesian")


freq_df <- broom::tidy(fit_lm, conf.int = TRUE) |>
  dplyr::select(term, estimate, conf.low, conf.high) |>
  dplyr::mutate(model = "Frequentist")


truth_df <- tibble::tibble(
  term = c("(Intercept)", paste0("x", 1:p)),
  truth = c(0, beta_true),
  model = "Truth"
)
```


```{r plot-results1}
ggplot2::ggplot() +

  # Bayesian posterior intervals
  ggdist::stat_pointinterval(
    data = bayes_draws,
    aes(x = estimate, y = term),
    .width = c(.5, .8, .95)
  ) +

  # Frequentist estimates + CI
  ggdist::geom_pointinterval(
    data = freq_df,
    aes(
      x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = term
    ),
    color = "red"
  ) +

  # Truth
  ggplot2::geom_point(
    data = truth_df,
    ggplot2::aes(x = truth, y = term),
    color = "blue",
    size = 3,
    shape = 4,
    stroke = 1.5
  ) +

  ggplot2::facet_wrap(~model) +

  ggplot2::theme_bw()
```


```{r plot-results2}
#| output: false
#| echo: false

## other approach
bayes_summary <- bayes_draws |>
  dplyr::group_by(term) |>
  tidybayes::median_qi(estimate, .width = 0.95) |>
  dplyr::transmute(
    term,
    estimate,
    conf.low = .lower,
    conf.high = .upper,
    model = "Bayesian"
  )


freq_plot <- freq_df |>
  dplyr::transmute(
    term,
    estimate,
    conf.low,
    conf.high,
    model = "Frequentist"
  )

plot_df <- dplyr::bind_rows(
  bayes_summary,
  freq_plot
)

ggplot2::ggplot(
  plot_df,
  ggplot2::aes(
    x = estimate,
    y = term,
    color = model
  )
) +

  ggdist::geom_pointinterval(
    ggplot2::aes(
      xmin = conf.low,
      xmax = conf.high
    ),
    position = ggplot2::position_dodge(width = 0.5)
  ) +

  ggplot2::geom_point(
    data = truth_df,
    ggplot2::aes(x = truth, y = term),
    inherit.aes = FALSE,
    color = "blue",
    shape = 4,
    size = 3,
    stroke = 1.2
  ) +

  ggplot2::geom_vline(
    xintercept = 0,
    linetype = 2
  ) +

  ggplot2::theme_bw()
```


## Visualizing the model:

```{r classic-bayes-plot}
# -----------------------------
# Prior-only model
# -----------------------------

fit_prior <- stan_glm(
  form,
  data = dat,
  prior = normal(0, 1.5),
  prior_intercept = normal(0, 0.25),
  chains = 4,
  iter = 2000,
  refresh = 0,
  prior_PD = TRUE
)

# -----------------------------
# Extract prior/posterior draws
# -----------------------------

coef_name <- "x1"  # coefficient of interest

post_draws <- posterior::as_draws_df(fit_bayes) |>
  dplyr::transmute(b = .data[[coef_name]])

prior_draws <- posterior::as_draws_df(fit_prior) |>
  dplyr::transmute(b = .data[[coef_name]])

draws <- dplyr::bind_rows(
  prior = prior_draws,
  posterior = post_draws,
  .id = "dist"
)

# -----------------------------
# Approximate likelihood
# -----------------------------

bhat <- coef(summary(fit_lm))[coef_name, "Estimate"]
se   <- coef(summary(fit_lm))[coef_name, "Std. Error"]

likelihood_data <- tibble::tibble(
  b = seq(bhat - 4 * se, bhat + 4 * se, length.out = 500)
) |>
  dplyr::mutate(
    density = dnorm(b, mean = bhat, sd = se),
    dist = "likelihood"
  )

# -----------------------------
# Plot
# -----------------------------

library(ggplot2)

ggplot(draws, aes(x = b, fill = dist)) +
  ggdist::stat_slab(alpha = 0.5) +
  geom_area(
    data = likelihood_data,
    aes(x = b, y = density, fill = "likelihood"),
    inherit.aes = FALSE,
    alpha = 0.5
  ) +
  scale_fill_manual(
    values = c(
      setNames(
        viridisLite::viridis(length(unique(draws$dist)),
                             option = "D", end = .85),
        unique(draws$dist)
      ),
      likelihood = "#CCBB44"
    )
  ) +
  theme_linedraw()
```


# Exercises: Bayesian updating and posterior probabilities

Use the [shiny app](https://shiny.rit.albany.edu/stat/binomial/) to plot your binomial distribution.
























